---
title: "ECON807 PS5: Business cycle properties in Canada"
author: "Daniel Sánchez-Pazmiño"
format: pdf
knitr:
  opts_chunk: 
    echo: true
    message: false
    warning: false
---


In this document I walk through my code, equations and results for the ECON807 PS5.

# Preliminaries

```{r}
#| label: setup

# Load libraries

library(tidyverse)
library(lubridate)
library(mFilter)
library(dynlm)

# Load the data

raw_national_accounts <- 
  read.csv('data.csv')
 
```
# Preparing the data

In this part, I prepare the dataset to present a time series of the various national accounts. 

```{r}
#| label: data-prep

# Transform to a date

df <- 
  raw_national_accounts %>% 
  mutate(
    date = ymd(paste(REF_DATE, '01')),
    month_year = floor_date(date, 'month')
  ) %>% 
  select(-REF_DATE) %>% 
  relocate(date, month_year) %>% 
  rename(account = 'Estimates',
         value = 'VALUE')

# Create a joint dataframe to plot all of my timeseries. Will be needing to create a lot of new dataframes

output <-
  df %>% 
  filter(account == 'Gross domestic product at market prices') %>% 
  select(month_year, account, value) %>% 
  mutate(account = 'Output')

consumption <-
  df %>% 
  filter(account %in% c('Household final consumption expenditure', 'Durable goods')) %>% 
  select(month_year, account, value) %>% 
  group_by(month_year) %>% 
  summarise(value = value[account == 'Household final consumption expenditure'] -
              value[account == 'Durable goods']) %>% 
  mutate(account = 'Consumption')

investment <-
  df %>% 
  filter(account %in% c('Gross fixed capital formation', 
                        'Durable goods', 
                        'General governments gross fixed capital formation')) %>% 
  select(month_year, account, value) %>% 
  group_by(month_year) %>% 
  summarise(value = 
              value[account == 'Gross fixed capital formation'] +
              value[account == 'Durable goods'] -
              value[account == 'General governments gross fixed capital formation'] )%>% 
  mutate(account = 'Investment')

government <-
  df %>% 
  filter(account %in% c ('General governments final consumption expenditure', 
                        'General governments gross fixed capital formation')) %>% 
  select(month_year, account, value) %>% 
  group_by(month_year) %>% 
  summarise(value =
              value[account == 'General governments final consumption expenditure'] +
              value[account == 'General governments gross fixed capital formation'] )%>% 
  mutate(account = 'Government Purchases')

time_series_graph <-
  output %>% 
  bind_rows(consumption) %>% 
  bind_rows(investment) %>% 
  bind_rows(government)

```

# (a) Presenting the national account time series

I present graphs with the time series below.

```{r}
#| label: time-series-graph
#| fig.height: 5
#| fig.width: 6

time_series_graph %>% 
  ggplot(aes(month_year, value/1000, colour = account)) + 
  geom_line(linewidth = 1) +
  theme_bw()+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = c('0.2', '0.8'))+
  labs(x = 'Time (t)',
       y = 'Value in billions of chained (2012) dollars ',
       colour = 'National Account',
       title = 'Time Series Graph of Select Canadian National Accounts',
       subtitle = 'Quarterly 1961-2022')

```
# (b) Some transformations

Below, I do the required transformations on the data.

```{r}
#| label: transformations

# (i) Log

time_series_transformed <-
  time_series_graph %>% 
  mutate(log = log(value))

# (ii) Baxter King filter

bk_log <- bkfilter(time_series_transformed$log, pl = 6, pu = 32, nfix = 12)

# Add to the dataframe

time_series_transformed <-
  time_series_transformed %>% 
  mutate(bk_log_trend = bk_log$trend,
         bk_log_cycle = bk_log$cycle)

# (iii) Hamilton (2016)

# Need to run a regression with lags of the dependent variable. 
# Must use a ts object for all of the different national accounts

output_ts <- ts(
  output,
  start = c(1961,1),
  frequency = 4
)

consumption_ts <- ts(
  consumption,
  start = c(1961,1),
  frequency = 4
)

investment_ts <-ts(
  investment,
  start = c(1961,1),
  frequency = 4
)

government_ts <- ts(
  government,
  start = c(1961,1),
  frequency = 4
)

# Run the regressions for all of the national accounts as specified
# using the dynlm package
# Include the residuals in each of the dataframes to be graphed

output_reg <- dynlm(
  value ~ L(value, 8) + L(value, 9) + L(value, 10) + L(value, 11),
  data = output_ts
)

residuals <- output_reg$residuals

output_ts <-
  output_ts %>% 
  cbind('residuals' = residuals)

```



